Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-04 source/sec_overparameterized.tex
minor updates
minor fixes
minor fix

improved convergence rate over FedAvg. Detailed proofs are deferred to Appendix Section~\ref{sec:interpolation}.

@@

\subsection{Geometric Convergence of FedAvg in the Overparameterized Setting}
Recall the federated learning problem $\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
with $F_{k}(\mathbf{w})=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\ell(\mathbf{w};\mathbf{x}_{k}^{j})$.
In this section, we consider the standard Empirical Risk Minimization (ERM) setting where $\ell$
is non-negative, $l$-smooth, and convex, and as before, each $F_{k}(\mathbf{w})$ is $L$-smooth and $\mu$-strongly convex. Note that $l\geq L$. This
setup includes many important problems in practice. In the overparameterized
setting, there exists $\mathbf{w}^{\ast}\in\arg\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
such that $\ell(\mathbf{w}^{\ast};\mathbf{x}_{k}^{j})=0$ for all
$\mathbf{x}_{k}^{j}$. We first show that FedAvg achieves geometric convergence
with linear speedup in the number of workers. 
\begin{theorem}
	\label{thm:overparameterized_general}In the overparameterized setting,
	FedAvg with communication every $E$ iterations
	and constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})})$
	has geometric convergence
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\frac{L}{2}(1-\overline{\alpha})^{T}\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}=\mathcal{O}\left(L\exp(-\frac{\mu}{E}\frac{NT}{l\nu_{\max}+L(N-\nu_{\min})})\cdot\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}
%
\textbf{Linear speedup and Communication Complexity} The linear speedup factor is on the order of $\mathcal{O}(N/E)$ %\frac{N}{E(l\nu_{\max}+L(N-\nu_{\min}))}/\frac{1}{l\nu_{\max}+L(1-\nu_{\min})}\approx\frac{Nl}{E(l+L(N-1))}=
for $N\leq\mathcal{O}(\frac{l}{L})$, i.e. FedAvg with $N$ workers and communication
every $E$ iterations provides a geometric convergence speedup factor
of $\mathcal{O}(N/E)$, for $N\leq\mathcal{O}(\frac{l}{L})$. When $N$ is above
this threshold, however, the speedup is almost constant in the number
of workers. This matches the findings in \cite{ma2017power}. Our
result also illustrates that $E$ can be taken $\mathcal{O}(T^{\beta})$
for any $\beta<1$ to achieve geometric convergence, achieving better communication efficiency than
the standard FL setting.

@@

We now turn to quadratic problems and show that the bound in Theorem \ref{thm:overparameterized_general} can be improved to $\mathcal{O}(\exp(-\frac{N}{E\kappa_{1}}t))$. We then propose a variant of FedAvg that has provable acceleration over FedAvg with SGD updates. The local device objectives are now given by the sum of squares {\small$F_{k}(\mathbf{w})=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(\mathbf{w}^{T}\mathbf{x}_{k}^{j}-z_{k}^{j})^{2}$},
and there exists $\mathbf{w}^{\ast}$ such that $F(\mathbf{w}^{\ast})\equiv0$. Two notions of condition number are important in our results:$\kappa_1$ which is based on local Hessians, and $\tilde{\kappa}$, which is termed the statistical condition number~\cite{liu2018accelerating,jain2017accelerating}. For their detailed definitions, please refer to Appendix Section~\ref{sec:interpolation}. Here we use the fact $\tilde{\kappa} \leq \kappa_1$.

@@

\begin{theorem}
	\label{thm:overparameterized_quadratic}For the overparamterized quadratic
	problem, FedAvg with communication every $E$
	iterations with constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+\mu(N-\nu_{\min})})$
	has geometric convergence:
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\mathcal{O}\left(L(1-\frac{N}{E(\nu_{\max}\kappa_{1}+(N-\nu_{\min}))})^{T}\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}

When $N=\mathcal{O}(\kappa_{1})$, the convergence rate is $\mathcal{O}((1-\frac{N}{E\kappa_{1}})^{T})=\mathcal{O}(\exp(-\frac{NT}{E\kappa_{1}}))$,
which exhibits linear speedup in the number of workers, as well as
a $1/\kappa_{1}$ dependence on the condition number $\kappa_{1}$.
Inspired by \cite{liu2018accelerating}, we propose the \textbf{MaSS
	accelerated FedAvg algorithm} (FedMaSS):
\begin{align*}
\mathbf{w}_{t+1}^{k} & =\begin{cases}
\mathbf{u}_{t}^{k}-\eta_{1}^{k}\mathbf{g}_{t,k} & \text{if }t+1\notin\mathcal{I}_{E},\\
\sum_{k=1}^{N}p_{k}\left[\mathbf{u}_{t}^{k}-\eta_{1}^{k}\mathbf{g}_{t,k}\right] & \text{if }t+1\in\mathcal{I}_{E},
\end{cases}\\
\mathbf{u}_{t+1}^{k} & =\mathbf{w}_{t+1}^{k}+\gamma^{k}(\mathbf{w}_{t+1}^{k}-\mathbf{w}_{t}^{k})+\eta_{2}^{k}\mathbf{g}_{t,k}. 
\end{align*}
When $\eta_{2}^{k}\equiv0$, this algorithm reduces to the Nesterov
accelerated FedAvg algorithm. In the next theorem, we demonstrate
that FedMaSS improves the convergence to $\mathcal{O}(\exp(-\frac{NT}{E\sqrt{\kappa_{1}\tilde{\kappa}}}))$.
To our knowledge, this is the first accelerated result of
FedAvg with momentum updates over SGD updates.
\begin{theorem}
	\label{thm:overparameterized_MaSS}For the overparamterized quadratic
	problem, FedMaSS with communication every $E$ iterations and constant
	step sizes $\overline{\eta}_1=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+\mu(N-\nu_{\min})}),\overline{\eta}_{2}=\frac{\overline{\eta}_{1}(1-\frac{1}{\tilde{\kappa}})}{1+\frac{1}{\sqrt{\kappa_{1}\tilde{\kappa}}}},\overline{\gamma}=\frac{1-\frac{1}{\sqrt{\kappa_{1}\tilde{\kappa}}}}{1+\frac{1}{\sqrt{\kappa_{1}\tilde{\kappa}}}}$
	has geometric convergence:
	\begin{align*}
	\mathbb{E}F(\overline{\vw}_{T}) & \leq\mathcal{O}\left((1-\frac{1}{E}\frac{N}{l\nu_{\max}\sqrt{\kappa_{1}\tilde{\kappa}}+(N-\nu_{\min})})^{T}\|w_{0}-w^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}
\textbf{Speedup of FedMaSS over FedAvg} 
To better understand the significance of the above result, we briefly discuss related works on accelerating SGD.
Nesterov and Heavy Ball updates are known to fail to accelerate over SGD in both the overparameterized
and convex settings~\cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}.
Thus in general one cannot hope to obtain acceleration results for
the FedAvg algorithm with Nesterov and Heavy Ball updates. 
Luckily, recent works in SGD~\cite{jain2017accelerating,liu2018accelerating} introduced
an additional compensation term to the Nesterov updates to address the non-acceleration issue. Surprisingly, 
we show the same approach can effectively improve the rate of FedAvg.
Comparing the convergence rate of FedMass (Theorem~\ref{thm:overparameterized_MaSS}) and FedAvg (Theorem~\ref{thm:overparameterized_quadratic}), when $N=\mathcal{O}(\sqrt{\kappa_{1}\tilde{\kappa}})$,
the convergence rate is $\mathcal{O}((1-\frac{N}{E\sqrt{\kappa_{1}\tilde{\kappa}}})^{T})=\mathcal{O}(\exp(-\frac{NT}{E\sqrt{\kappa_{1}\tilde{\kappa}}}))$
as opposed to $\mathcal{O}(\exp(-\frac{NT}{E\kappa_{1}}))$. Since
$\kappa_{1}\geq\tilde{\kappa}$, this implies a speedup factor of
$\sqrt{\frac{\kappa_{1}}{\tilde{\kappa}}}$ for FedMaSS. On the other
hand, the same linear speedup in the number of workers holds for $N$
in a smaller range of values.

>>>>

improved convergence rate over FedAvg. Detailed proofs are deferred to Appendix Section~\ref{sec:interpolation}. In particular, we do not need Assumptions~\ref{ass:boundedvariance} and~\ref{ass:subgrad2} and use modified versions of Assumptions~\ref{ass:lsmooth} and~\ref{ass:stroncvx} detailed in this section.

@@

\subsection{Geometric Convergence of FedAvg in the Overparameterized Setting}
Recall the federated learning problem $\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
with $F_{k}(\mathbf{w})=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\ell(\mathbf{w};\mathbf{x}_{k}^{j})$.
In this section, we consider the standard Empirical Risk Minimization (ERM) setting where $\ell$
is non-negative, $l$-smooth, and convex, and as before, each $F_{k}(\mathbf{w})$ is $L$-smooth and $\mu$-strongly convex. Note that $l\geq L$. This
setup includes many important problems in practice. In the overparameterized
setting, there exists $\mathbf{w}^{\ast}\in\arg\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
such that $\ell(\mathbf{w}^{\ast};\mathbf{x}_{k}^{j})=0$ for all
$\mathbf{x}_{k}^{j}$. We first show that FedAvg achieves geometric convergence
with linear speedup in the number of workers. 
\begin{theorem}
	\label{thm:overparameterized_general}In the overparameterized setting,
	FedAvg with communication every $E$ iterations
	and constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})})$
	has geometric convergence
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\frac{L}{2}(1-\overline{\alpha})^{T}\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}=\mathcal{O}\left(L\exp\left(-\frac{\mu}{E}\frac{NT}{l\nu_{\max}+L(N-\nu_{\min})}\right)\cdot\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}
%
\textbf{Linear speedup and Communication Complexity} The linear speedup factor is on the order of $\mathcal{O}(N/E)$ %\frac{N}{E(l\nu_{\max}+L(N-\nu_{\min}))}/\frac{1}{l\nu_{\max}+L(1-\nu_{\min})}\approx\frac{Nl}{E(l+L(N-1))}=
for $N\leq\mathcal{O}(\frac{l}{L})$, i.e. FedAvg with $N$ workers and communication
every $E$ iterations provides a geometric convergence speedup factor
of $\mathcal{O}(N/E)$, for $N\leq\mathcal{O}(\frac{l}{L})$. When $N$ is above
this threshold, however, the speedup is almost constant in the number
of workers. This matches the findings in \cite{ma2017power}. Our
result also illustrates that $E$ can be taken $\mathcal{O}(T^{\beta})$
for any $\beta<1$ to achieve geometric convergence, achieving better communication efficiency than
the standard FL setting.

@@

We now turn to quadratic problems and show that the bound in Theorem \ref{thm:overparameterized_general} can be improved to $\mathcal{O}(\exp(-\frac{N}{E\kappa_{1}}t))$. We then propose a variant of FedAvg that has provable acceleration over FedAvg with SGD updates. The local device objectives are now given by the sum of squares {\small$F_{k}(\mathbf{w})=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(\mathbf{w}^{T}\mathbf{x}_{k}^{j}-z_{k}^{j})^{2}$},
and there exists $\mathbf{w}^{\ast}$ such that $F(\mathbf{w}^{\ast})\equiv0$. Two notions of condition number are important in our results:$\kappa_1$ which is based on local Hessians, and $\tilde{\kappa}$, which is termed the statistical condition number~\cite{liu2018accelerating,jain2017accelerating}. For their detailed definitions, please refer to Appendix Section~\ref{sec:interpolation}. Here we use the fact $\tilde{\kappa} \leq \kappa_1$. Recall $\nu_{\max}=\max _k p_k N$ and $\nu_{\min}=\min _k p_k N$.

@@

\begin{theorem}
	\label{thm:overparameterized_quadratic}For the overparamterized quadratic
	problem, FedAvg with communication every $E$
	iterations with constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+\mu(N-\nu_{\min})})$
	has geometric convergence:
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\mathcal{O}\left(L\exp(-\frac{NT}{E(\nu_{\max}\kappa_{1}+(N-\nu_{\min}))})\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}

When $N=\mathcal{O}(\kappa_{1})$, the convergence rate is $\mathcal{O}((1-\frac{N}{E\kappa_{1}})^{T})=\mathcal{O}(\exp(-\frac{NT}{E\kappa_{1}}))$,
which exhibits linear speedup in the number of workers, as well as
a $1/\kappa_{1}$ dependence on the condition number $\kappa_{1}$.
Inspired by \cite{liu2018accelerating}, we propose the \textbf{MaSS
	accelerated FedAvg algorithm} (FedMaSS):
\begin{align*}
\mathbf{w}_{t+1}^{k} & =\begin{cases}
\mathbf{u}_{t}^{k}-\eta_{1}^{k}\mathbf{g}_{t,k} & \text{if }t+1\notin\mathcal{I}_{E},\\
\sum_{k \in \cS_{t+1}}\left[\mathbf{u}_{t}^{k}-\eta_{1}^{k}\mathbf{g}_{t,k}\right] & \text{if }t+1\in\mathcal{I}_{E},
\end{cases}\\
\mathbf{u}_{t+1}^{k} & =\mathbf{w}_{t+1}^{k}+\gamma^{k}(\mathbf{w}_{t+1}^{k}-\mathbf{w}_{t}^{k})+\eta_{2}^{k}\mathbf{g}_{t,k}. 
\end{align*}
When $\eta_{2}^{k}\equiv0$, this algorithm reduces to the Nesterov
accelerated FedAvg algorithm. In the next theorem, we demonstrate
that FedMaSS improves the convergence to $\mathcal{O}(\exp(-\frac{NT}{E\sqrt{\kappa_{1}\tilde{\kappa}}}))$.
To our knowledge, this is the first acceleration result of
FedAvg with momentum updates over SGD updates.
\begin{theorem}
	\label{thm:overparameterized_MaSS}For the overparamterized quadratic
	problem, FedMaSS with communication every $E$ iterations and constant
	step sizes $\overline{\eta}_1=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+\mu(N-\nu_{\min})}),\overline{\eta}_{2}=\frac{\overline{\eta}_{1}(1-\frac{1}{\tilde{\kappa}})}{1+\frac{1}{\sqrt{\kappa_{1}\tilde{\kappa}}}},\overline{\gamma}=\frac{1-\frac{1}{\sqrt{\kappa_{1}\tilde{\kappa}}}}{1+\frac{1}{\sqrt{\kappa_{1}\tilde{\kappa}}}}$
	has geometric convergence:
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\mathcal{O}\left(L\exp(-\frac{NT}{E(\nu_{\max}\sqrt{\kappa_{1}\tilde{\kappa}}+(N-\nu_{\min}))})\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}
\textbf{Speedup of FedMaSS over FedAvg} 
To better understand the significance of the above result, we briefly discuss related works on accelerating SGD.
Nesterov and Heavy Ball updates are known to fail to accelerate over SGD in both the overparameterized
and convex settings~\cite{liu2018accelerating,kidambi2018insufficiency,liu2018toward,yuan2016influence}.
Thus in general one cannot hope to obtain acceleration results for
the FedAvg algorithm with Nesterov and Heavy Ball updates. 
Luckily, recent works in SGD~\cite{jain2017accelerating,liu2018accelerating} introduced
an additional compensation term to the Nesterov updates to address the non-acceleration issue. Surprisingly, 
we show the same approach can effectively improve the rate of FedAvg.
Comparing the convergence rate of FedMass (Theorem~\ref{thm:overparameterized_MaSS}) and FedAvg (Theorem~\ref{thm:overparameterized_quadratic}), when $N=\mathcal{O}(\sqrt{\kappa_{1}\tilde{\kappa}})$,
the convergence rate is $\mathcal{O}((1-\frac{N}{E\sqrt{\kappa_{1}\tilde{\kappa}}})^{T})=\mathcal{O}(\exp(-\frac{NT}{E\sqrt{\kappa_{1}\tilde{\kappa}}}))$
as opposed to $\mathcal{O}(\exp(-\frac{NT}{E\kappa_{1}}))$. Since
$\kappa_{1}\geq\tilde{\kappa}$, this implies a speedup factor of
$\sqrt{\frac{\kappa_{1}}{\tilde{\kappa}}}$ for FedMaSS. On the other
hand, the same linear speedup in the number of workers holds for $N$
in a smaller range of values.
